---
title: "Normal posterior approximation - iid Poisson model"
author: "Mattias Villani"
editor: visual
---

```{r}
#| echo: false
#| output: false
prettycols = c("#6C8EBF", "#c0a34d", "#780000","#007878","#B5C6DF","#EADAAA","#AE6666")
```

## Problem 5 - Normal approximation

The file `bugs.csv` contains a dataset with the number of bugs and some other explanatory variables for $n=91$ releases of several software projects. We will here do a Bayesian analysis using only the variable `nBugs`, which we will store in a vector `y`, for simplicity. Load the data like this:

```{r}
data = read.csv("https://github.com/mattiasvillani/BayesLearnCourse/raw/master/assignment/bugs.csv", 
                header = TRUE)
y = data$nBugs # number of bugs, a vector with n = 91 observations
```

You can ignore that some of the observations actually comes from the same project at different releases, and assume that the observations are *independent* and identically distributed. Consider first the model

$$Y_1,Y_2,\ldots,Y_n \vert \lambda \overset{iid}{\sim}\mathrm{Poisson}(\lambda),$$

where the parameter $\lambda = \mathbb{E}(Y)$ must be positive. Our interest is in obtaining the posterior distribution for $\lambda$. If we use the conjugate Gamma prior we can obtain the posterior in closed form (it is also Gamma). We will however use a **normal posterior approximation** here, since next problem will extend the model to a Poisson regression model with covariates, and for that model there is no conjugate prior.

A normal approximation of the posterior for $\lambda$ is not a great idea, however; a normal approximation can give probability mass on negative values for $\lambda$, and we need to have $\lambda>0$ in the Poisson model. We therefore instead parameterize the Poisson model as

$$Y_1,Y_2,\ldots,Y_n \vert \theta \overset{iid}{\sim}\mathrm{Poisson}(\exp(\theta))$$

where $\theta$ can be any value on the real line $-\infty < \theta <\infty$ because the exponential function makes sure that $\lambda = \exp(\theta) >0$. This kind of reparameterization usually makes the posterior more normal and the approximation more accurate. An additional benefit of working the $\theta$ is that we can use *unconstrained* *maximization* without specifying a \texttt{lower} bound in the optimizer.

A Normal posterior approximation for $\theta$ implies a LogNormal approximation of the posterior for $\lambda = \exp(\theta)$, which therefore *guarantees that all posterior probability mass is on positive values of* $\lambda$.

## LogNormal distribution

Recall the definition of a LogNormal distribution:

If $X \sim N(\mu\, \sigma^2)$ then $Y=\exp(X)\sim \mathrm{LogNormal}(\mu, \sigma^2)$.

Use the prior $\theta \sim N(\log(4), 0.5^2)$ which implies a LogNormal prior for $\lambda$. The prior in both parameterizations are plotted below.

```{r}
nSim = 10000
thetaDraws = rnorm(nSim, mean = log(4), sd = 0.5)
lambdaDraws = exp(thetaDraws)
par(mfrow = c(1,2))
hist(thetaDraws, 50, xlab = expression(theta), ylab = "prior density", col = prettycols[5], main = paste("prior for", expression(theta)))
hist(lambdaDraws, 50, xlab = expression(lambda), ylab = "prior density", col = prettycols[6], main = paste("prior for", expression(lambda)))
```

#### Problem 5a)

Do a normal approximation of the posterior distribution of $\theta$ using optimization and plot that normal approximation. Use this posterior approximation for $\theta$ to simulate the approximate posterior of $\lambda = \exp(\theta)$.

#### [Solution 5a)]{style="color:red;"}

We start by setting up the (unnormalized) log posterior by coding up the log-likelihood and the log-prior and finally adding them up:

```{r}
LogPostPoisson <- function(theta, y, mu0, tau0){
  logLik = sum(dpois(y, exp(theta), log = TRUE))
  logPrior = dnorm(theta, mu0, tau0, log = TRUE)
  return(logLik + logPrior)
}
```

We use the `optim` function to maximize the log posterior (with the argument `control=list(fnscale=-1)` that multiplies the objective function by $-1$ to turn the problem into a minimization problem, which is what `optim` handles). The initial value is set to $0$ for simplicity ( $\log(\bar y)$ would have been better).

```{r}
mu0 = log(4)
tau0 = 0.5
initVal <- 0; 
OptimResults<-optim(initVal, LogPostPoisson, gr=NULL, y, mu0, tau0,
  method = c("BFGS"), control=list(fnscale=-1), hessian=TRUE)
postMode = OptimResults$par[1]
postVar = -1/OptimResults$hessian[1] # inv(J) - Approx posterior variance
postStd <- sqrt(postVar)          # Approximate stdev
message("The posterior mode is ", round(postMode, 3), " and the approximate posterior standard deviation is ", round(postStd, 3), ".") 
```

Hence the approximate posterior is

$$
\theta \vert \boldsymbol{y} \overset{\mathrm{approx}}{\sim} N(`r round(postMode,3)`, `r round(postStd,3)`)
$$

```{r}
thetaGrid = seq(postMode - 4*postStd, postMode + 4* postStd, length = 1000)
plot(thetaGrid, dnorm(thetaGrid, postMode, postStd), type = "l", lwd = 3, 
     xlab = expression(theta), ylab = "posterior density", col = prettycols[3])
```

#### Problem 5b)

Use the posterior approximation for $\theta$ in Problem 5a) to simulate from the approximate posterior of $\lambda = \exp(\theta)$. Plot a histogram.

#### [Solution 5b)]{style="color:red;"}

```{r}
nSim = 10000
thetaDraws = rnorm(nSim, postMode, postStd)
lambdaDraws = exp(thetaDraws)
hist(lambdaDraws, 50, freq = FALSE, col = prettycols[6],
     xlab = expression(lambda), ylab = "posterior density", main = "Posterior lambda")
```

This posterior is slightly skewed (we actually know it is logNormal).
