---
title: "Solution to computer exam in Bayesian learning"
author: "Per Sidén"
date: "2019-06-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results="markup")
```

First load all the data into memory by running the R-file given at the exam
```{r}
rm(list=ls())
source("ExamData.R")
set.seed(1)
```

## Problem 1

### 1a
```{r}
nSamples = 10000 # 100000 #
beta = 20
postSamples1 = rbeta(nSamples,sqrt(4),beta)
postSamples2 = rbeta(nSamples,sqrt(16),beta)
par(mfrow=c(1,2))
hist(postSamples1,30,freq=F, main="Posterior based on\n samples for c=4",xlab="theta")
hist(postSamples2,30,freq=F, main="Posterior based on\n samples for c=16",xlab="theta")
```

### 1b

```{r}
print(c(mean(postSamples1>.1),mean(postSamples2>.1)))
```
Answer: The probabilities are roughly 0.36 and 0.81.

### 1c
Find the value of c that gives the maximum expected utility
```{r}
cgrid = seq(4,20,.5)
utility <- function(theta,c){
  return(100+20*log(theta)-c)
}
expectedUtility <- function(c){
  postSamples = rbeta(nSamples,sqrt(c),beta)
  return(mean(sapply(postSamples,utility,c=c)))
}
EU = sapply(cgrid,expectedUtility)
cOpt = cgrid[which.max(EU)]
par(mfrow=c(1,1))
plot(cgrid,EU,type="l",ylab="Expected utility",xlab="c")
points(x=cOpt, y = max(EU), col = "red")
print(cOpt)
```
Answer: The optimal value is roughly at $c=10.5$ (but can be between 9 and 11.5 due to randomness).

```{r}
# # Incorrect solution: evaluate the utility at the mean, gives c=8.5
# thetaMean <- function(c){
#   return(sqrt(c)/(sqrt(c)+beta))
# }
# 
# meanU = mapply(utility,thetaMean(cgrid),cgrid)
# cOptMean = cgrid[which.max(meanU)]
# plot(cgrid,meanU,type="l")
# points(x=cOptMean, y = max(meanU), col = "red")
# print(cOptMean)
```

## Problem 2

### 2a
```{r}
# Reading data
load(file = 'ebay.RData') 

gridn = 1000
thetaGrid <- seq(0,1,length = gridn)
logPost <- function(theta,x){
  
  # evaluating the log-likelihood                                    
  logLik <- sum(dbinom(x,size=50,prob=theta,log=T))
  # logLik <- sum(x*log(theta) + (50-x)*log(1-theta))
  
  # evaluating the prior
  logPrior <- 2*log(1-theta)
  
  # add the log prior and log-likelihood together to get log posterior
  logPost <- logLik + logPrior
  # if (abs(logPost) == Inf || is.na(logPost)) logPost = -20000;
  return(logPost)
}

logPostUnnorm <- vector(length=length(thetaGrid))
for(i in 1:length(thetaGrid)) logPostUnnorm[i] <- logPost(thetaGrid[i],ebay)
# logPostUnnorm = logPostUnnorm - max(logPostUnnorm) # For numerical stability
gridWidth = thetaGrid[2] - thetaGrid[1]
postGrid <- (1/gridWidth)*exp(logPostUnnorm)/sum(exp(logPostUnnorm))
plot(thetaGrid, postGrid, type = "l", lwd = 2, main="Posterior", 
     ylab = "Density", xlab = expression(theta))
thetaMode = thetaGrid[which.max(postGrid)]
print(thetaMode)
```
Answer: The mode is at $\theta=0.103$.

### 2b

We now use the GibbsMixPois function in the ExamData.R code to simulate with $K=2$ (Hide iteration count output using message=FALSE)
```{r,message=FALSE}
set.seed(100)
GibbsResults2 <- GibbsMixPois(x = ebay, nComp = 2, alpha = 1, alphaGamma = 1, betaGamma = 1, 
                              xGrid = xGrid, nIter = 500)
```

Plot trajectories and cumulative means
```{r}
par(mfrow=c(2,2))
plot(GibbsResults2$thetaSample[,1],type="l",xlab="Iteration",ylab="theta1",main="Trajectory")
plot(GibbsResults2$thetaSample[,2],type="l",xlab="Iteration",ylab="theta2",main="Trajectory")
plot(cumsum(GibbsResults2$thetaSample[,1])/seq(1:500),type="l",xlab="Iteration",
     ylab="theta1",main="Cumulative mean")
plot(cumsum(GibbsResults2$thetaSample[,2])/seq(1:500),type="l",xlab="Iteration",
     ylab="theta2",main="Cumulative mean")
```

A burn-in of 50 iterations is sufficient for the chain to reach its stationary distribution, which can be seen in the trajectory plots.

### 2c
```{r}
par(mfrow=c(1,1))
dataDistr = bidsCounts/sum(bidsCounts)

# thetaMode = .1 # If 2a not solved
binoDistr = dbinom(xGrid, size = 50, prob = thetaMode)
plot(xGrid, as.vector(dataDistr), type = "o", lwd = 3, col = "black", pch = 'o', cex = 0.6, 
     ylim = c(0,0.2), main = "Fitted models")
lines(xGrid, binoDistr, type = "o", lwd = 3, col = "red", pch = 'o', cex = 0.6)
lines(xGrid, GibbsResults2$mixDensMean, type = "o", lwd = 3, col = "purple", pch = 'o', cex = 0.6)
legend(x = 8, y = 0.2, legend = c("Data", "Binomial","Poisson mixture"), 
       col = c("black","red","purple"), lty = c(1,1), lwd = c(3,3), cex = 0.8)
```

```{r}

```


The binomial model fits the data quite badly, which can mainly be seen in that the number of zero-observations is much higher in the dataset, than what the density suggests. The Poisson mixture gives a better fit.

## Problem 3

See solution on paper.

## Problem 4

```{r}
# Load cellphone lifetime data
load(file = 'cellphones.RData')
x = cellphones
```

### 4a

```{r}

# hist(x)
n = length(x)
alpha1 = 2
beta1 = 1
alpha2 = 10
beta2 = 10

thetaGrid = seq(0,4,.01)
plot(thetaGrid,dgamma(thetaGrid,alpha1,beta1),type="l",ylab="",xlab="theta",main="Gamma prior densities")
lines(thetaGrid,dgamma(thetaGrid,alpha2,beta2),col="red")
legend(x=3,y=.3,c("M1","M2"),col = c("black","red"), lty = c(1,1))
```

The $M_1$ prior has mean $\frac21 = 2$ and variance $\frac{2}{1\cdot1} = 2$. The $M_2$ prior has mean $\frac{10}{10} = 1$ and variance $\frac{10}{10\cdot10} = 0.1$. Comparing the variances tells us that the $M_2$ prior is more informative and this can also be seen by plotting the densities. This is because the $M_2$ prior is more concentrated and has thinner tails.

### 4b
```{r}
logprior <- function(theta,alpha,beta){
  return(dgamma(theta,shape=alpha,rate=beta,log=T))
}
loglik <- function(x,theta){
  return(sum(dexp(x,theta,log=T)))
}
logposterior <- function(theta,alpha,beta,x){
  return(dgamma(theta,shape=alpha+length(x),rate=beta+sum(x),log=T))
}
logmarglik <- function(theta,alpha,beta,x){
  return(loglik(x,theta) + logprior(theta,alpha,beta) - logposterior(theta,alpha,beta,x))
}

lms = c(logmarglik(1,alpha1,beta1,x),logmarglik(1,alpha2,beta2,x))
unnormProbs = .5*exp(lms)
probs = unnormProbs/sum(unnormProbs)
print(probs)
```

The posterior probability of $M_1$ is 0.617 so this is more probable, however the data does not strongly suggest that any of the models is better than the other.

### 4c

The interval can be computed in several ways, but using simulation is simple
```{r}
ndraws = 100000
xTildeDraws = rep(0,ndraws)
for(i in 1:ndraws){
  M = rbinom(1,1,probs[2]) + 1 # Simulate which model to use
  if(M==1){
    theta = rgamma(1,shape=alpha1+length(x),rate=beta1+sum(x))
  } else {
    theta = rgamma(1,shape=alpha2+length(x),rate=beta2+sum(x))
  }
  xTildeDraws[i] = rexp(1,theta)
}

print(quantile(xTildeDraws,probs = c(.05,.95)))
```

The interval is roughly (0.12,7.5).
